# 📦 1. Load Required R Packages

This section loads all the libraries we will use throughout the pipeline. The `tidyverse` is used for general data wrangling and plotting; `LipidSigR` is the main lipidomics analysis tool; rgoslin is used for parsing lipid names; and other libraries support visualization, statistical tests, and data cleaning.

```{r}
# Core tidyverse: includes dplyr, ggplot2, readr, tibble, etc.
library(tidyverse)
library(purrr)

# Cleaning and input tools
library(janitor)       # For cleaning column names
library(readxl)        # To read .xlsx files

# Lipidomics tools
library(LipidSigR)     # Main analysis package
library(rgoslin)       # Lipid name parser

# Supporting packages
library(SummarizedExperiment)  # Core data container
library(ggrepel)               # Non-overlapping plot labels
library(broom)                 # Clean test output
library(car)                   # For Levene’s test (equal variances)
```

# 📥 2. Load and Clean Input Data

We load the lipidomics abundance matrix and the accompanying sample metadata, ensuring their sample names match.

```{r}
# Load metadata (sample information)
coldata <- read_xlsx("Edited copy of dt_lipidomics_praise[4].xlsx") %>%
  dplyr::mutate(`Sample ID` = gsub("-", "_", `Sample ID`))  # Ensure consistent sample names

# Load abundance matrix (lipid intensities)
matrix <- read_xlsx("Edited copy of praises_matrix.xlsx")

# Extract sample IDs
sample_ids <- coldata$`Sample ID`
matrix_sample_cols <- setdiff(names(matrix), c("LipidGroup", "Class"))

# Basic integrity checks
setdiff(sample_ids, matrix_sample_cols)    # Should return character(0)
setdiff(matrix_sample_cols, sample_ids)    # Should return character(0)
length(sample_ids) == length(matrix_sample_cols)  # Should be TRUE
all(sample_ids == matrix_sample_cols)             # Should be TRUE (optional)
```

Change Days of incubation to D0

# 🧪 3. Prepare Lipid Data for LipidSigR

Here we prepare the three required inputs to create a SummarizedExperiment object:

-   abundance: table of lipid intensities

-   group_info: metadata with group labels

-   goslin_annotation: lipid structure information from rgoslin

We also filter out any unparseable lipid names, which are required by LipidSigR’s internal tools.

```{r}
# 3.1 Format abundance table
abundance <- matrix %>%
  dplyr::rename(feature = LipidGroup)  # LipidSigR requires 'feature' as the name column

# 3.2 Parse lipid names into standardized format
parsed_lipids <- parseLipidNames(abundance$feature)

# 3.3 Create annotation table from parsed results
goslin_annotation <- parsed_lipids
```

# 🧾 4. Create Sample Metadata for LipidSigR

Here we format sample metadata into the structure LipidSigR expects:

-   sample_name: matches abundance matrix

-   group: biological condition (e.g. KO vs WT)

-   timepoint: custom column for future stratification

-   label_name, pair: required fields (pair = NA if unpaired design)

```{r}
# Define sample groups and timepoints
group_info <- coldata %>%
  dplyr::rename(sample_name = `Sample ID`) %>%
  mutate(
    group = case_when(
      str_detect(sample_name, "ROSA") ~ "WT",
      str_detect(sample_name, "SLC") ~ "KO",
      TRUE ~ "Unknown"
    ),
    timepoint = case_when(
      str_detect(sample_name, "^R_") ~ "Replete",
      str_detect(sample_name, "^D0_") ~ "Day0",
      str_detect(sample_name, "^D1_") ~ "Day1",
      str_detect(sample_name, "^D2_") ~ "Day2",
      str_detect(sample_name, "^D3_") ~ "Day3",
      str_detect(sample_name, "^D4_") ~ "Day4",
      TRUE ~ "Unknown"
    ),
    label_name = sample_name,
    pair = NA
  )

# Save full version of metadata (to restore timepoint info later)
group_info_full <- group_info

# Select required columns
group_info <- group_info %>%
  select(sample_name, label_name, group, pair)

# Match order of sample columns with abundance table
sample_names_abundance <- setdiff(names(abundance), c("feature", "Class"))
group_info <- group_info %>%
  arrange(factor(sample_name, levels = sample_names_abundance))

```

# 🧹 5. Filter Unrecognized Lipids and Build SummarizedExperiment

Many lipid names may not be parseable by rgoslin, and LipidSigR requires recognized lipids for structure-aware analyses. Here we:

Keep only recognized lipid species,

Build the SummarizedExperiment object for LipidSigR analysis,

Add timepoint metadata back to the object.

```{r}
# 5.1 Keep only recognized lipids
recognized_lipids <- parsed_lipids %>%
  filter(Grammar != "NOT_PARSEABLE") %>%
  pull(Original.Name)

abundance_filtered <- abundance %>%
  filter(feature %in% recognized_lipids) %>%
  select(-Class)  # Optional: drop unused column

goslin_annotation_filtered <- parsed_lipids %>%
  filter(Original.Name %in% abundance_filtered$feature,
         Grammar != "NOT_PARSEABLE")

# 5.2 Build SummarizedExperiment object
se <- as_summarized_experiment(
  abundance = abundance_filtered,
  goslin_annotation = goslin_annotation_filtered,
  group_info = group_info,
  se_type = "de_two",       # We will compare two groups: KO vs WT
  paired_sample = FALSE     # Samples are independent
)

# 5.3 Add timepoint metadata for downstream stratification
colData(se)$timepoint <- group_info_full$timepoint
```

# 📊 6. Exploratory Analysis with PCA and Sample Similarity

In this section, we apply unsupervised methods to explore the dataset before conducting any hypothesis testing. These approaches help us:

Detect group clustering (e.g., KO vs WT),

Identify outlier samples,

Understand variance structure in the dataset.

## 🔹 6.1 Principal Component Analysis (PCA)

PCA reduces the complexity of high-dimensional data (many lipid species) into a few new dimensions (PC1, PC2, etc.) that capture the most variance in the dataset. It helps answer:

Are samples with similar biological labels grouping together?

Which lipid features drive the separation?

We also use k-means clustering on the PCA scores to check for data-driven groupings.

```{r}
# 🧱 1. Construct SummarizedExperiment
processed_se <- as_summarized_experiment(
  abundance = abundance_filtered,
  goslin_annotation = goslin_annotation_filtered,
  group_info = group_info,
  se_type = "de_two",
  paired_sample = FALSE
)

# 🧬 2. Add back timepoint info (from full metadata)
colData(processed_se)$sample_name <- rownames(colData(processed_se))
colData(processed_se) <- colData(processed_se) %>%
  as.data.frame() %>%
  left_join(group_info_full %>% select(sample_name, timepoint), by = "sample_name") %>%
  DataFrame()

# 🔄 3. Process data using LipidSigR::data_process() — REQUIRED for PCA and profiling
processed_se <- data_process(
  se = processed_se,
  exclude_missing = TRUE,
  exclude_missing_pct = 70,
  replace_na_method = "min",
  replace_na_method_ref = 0.5,
  normalization = "Percentage",
  transform = "log10"
)

result_pca <- dr_pca(
  processed_se,
  scaling = TRUE,
  centering = TRUE,
  clustering = 'kmeans',
  cluster_num = 2,
  feature_contrib_pc = c(1, 2),
  plot_topN = 10
)


# Plot outputs:
result_pca$static_pca                 # Sample-level PCA plot
result_pca$static_screePlot           # Variance explained per PC
result_pca$static_feature_contribution # Correlation circle (lipids vs PCs)
result_pca$static_variablePlot        # Barplot of top lipids driving variance
```

## 🔥 6.2 Sample-Sample Correlation Heatmap

This heatmap shows the pairwise similarity of samples based on lipid abundance profiles.

Pearson correlation is used to quantify how similar each pair of samples is.

Hierarchical clustering groups samples that behave similarly.

This is helpful to spot batch effects, replicates, or unexpected groupings.

```{r}
# Generate a correlation heatmap across all samples
result_heatmap <- heatmap_correlation(
  processed_se,
  char = NULL,                 # No lipid grouping yet — raw sample correlation
  transform = 'log10',         # Apply log10 transform before correlation
  correlation = 'pearson',     # Use Pearson correlation
  distfun = 'maximum',         # Distance metric for clustering
  hclustfun = 'average',       # Clustering method
  type = 'sample'              # Compute correlation between samples
)

# Visualize the sample-to-sample correlation heatmap
result_heatmap$static_heatmap
```

# 🧬 7. Lipid Class Composition Across Samples

Now we explore lipid composition by lipid class, which gives a broader view of lipid metabolism rather than focusing on individual species.

This profiling step:

Aggregates all lipids within each class (e.g., PC, PE, TG),

Computes relative abundance of each class per sample,

Helps identify class-wide shifts in lipid composition.

```{r}
# Group lipid data by class and plot overall composition
result_lipid <- lipid_profiling(
  processed_se,
  char = "class"  # Aggregate lipids by class
)

# Visualization outputs:
result_lipid$static_char_barPlot          # Bar plot of class abundance per sample
result_lipid$static_lipid_composition     # Stacked bar plot showing relative class proportions

```

# 🧪 8. Manual Normalization, Transformation, and PCA (Without LipidSigR)

This section replicates core preprocessing steps manually, providing transparency into what LipidSigR is doing "under the hood." We normalize the data, log-transform it, and perform PCA directly.

## 🔹 8.1 Manual Normalization and Log Transformation

We:

Remove annotation columns,

Impute missing values with half of the smallest non-zero value,

Normalize lipid intensities across samples to sum to 100%,

Apply a log10 transform for stabilizing variance.

```{r}
# 1. Subset to lipid values only
data <- matrix %>%
  select(-LipidGroup, -Class)

# 2. Impute missing values with half the minimum observed
min_val <- min(data, na.rm = TRUE)
data[is.na(data)] <- 0.5 * min_val

# 3. Restore lipid names as rownames
rownames(data) <- matrix$LipidGroup

# 4. Normalize each column (sample) to sum to 100
data_norm <- sweep(data, 2, colSums(data), FUN = "/") * 100

# 5. Log10-transform (adding pseudocount to avoid -Inf)
data_log <- log10(data_norm + 1e-6)
```

## 🔹 8.2 Correlation Heatmap (Manual)

This plot shows how similar each sample is to others based on lipid abundance. Pearson correlation is commonly used for this purpose.

```{r}
# Calculate correlation matrix using log-transformed data
cor_matrix <- cor(data_log, use = "pairwise.complete.obs")

# Create detailed sample names
detailed_names <- paste(coldata$`Cell type`, coldata$`FATP2 status`, coldata$`Days in incubation`, sep = "_")

# Plot correlation heatmap
pheatmap::pheatmap(
  cor_matrix,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  main = "Sample Correlation Matrix",
  labels_row = detailed_names,
  labels_col = detailed_names,
  fontsize_row = 9,
  fontsize_col = 9,
  fontsize = 11,
  display_numbers = FALSE,
  number_format = "%.2f",
  color = colorRampPalette(c("navy", "white", "firebrick3"))(50)
)
```

## PCA

```{r}
# 1. Run PCA (transpose because rows = lipids, columns = samples)
pca <- prcomp(t(data_log), center = TRUE, scale. = TRUE)

# 2. Extract PC1 and PC2 scores per sample
pca_scores <- as.data.frame(pca$x[, 1:2])
pca_scores$SampleID <- rownames(pca_scores)

# 3. Merge PCA scores with metadata
pca_scores_annotated <- pca_scores %>%
  left_join(
    coldata %>% dplyr::rename(SampleID = `Sample ID`),
    by = "SampleID"
  )

# 4. Plot PCA with annotations
ggplot(pca_scores_annotated, aes(x = PC1, y = PC2,
                                 color = `Days in incubation`,
                                 shape = `FATP2 status`)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Lipid Data (Manual)",
    x = paste0("PC1 (", round(100 * summary(pca)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC2 (", round(100 * summary(pca)$importance[2, 2], 1), "% variance)")
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

```

## 🧬 9. Lipid Class Composition by Sample

This section aggregates lipid abundance **by class** and visualizes their relative distribution across samples. This helps assess broad trends (e.g., enrichment of TAGs, PCs, etc.) before diving into specific lipid species.

```{r}
class_abundance <- matrix %>%
  pivot_longer(cols = -c(LipidGroup, Class), names_to = "Sample", values_to = "Abundance") %>%
  mutate(Abundance = replace_na(Abundance, 0)) %>%
  group_by(Sample, Class) %>%
  summarise(Total = sum(Abundance), .groups = "drop") %>%
  group_by(Sample) %>%
  mutate(Relative = Total / sum(Total) * 100) %>%
  ungroup()

# Horizontal stacked barplot
ggplot(class_abundance, aes(x = Sample, y = Relative, fill = Class)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Lipid Class Composition by Sample", x = "Sample", y = "Relative Abundance (%)") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8), axis.text.x = element_text(size = 10), legend.position = "right")
```

## Tru

## 📊 10. Distribution Testing for Choosing DE Method

⚠🛑📈🧾 

To determine whether to apply a **parametric t-test** or **non-parametric Wilcoxon test**, we assess: - **Normality** using Shapiro-Wilk - **Variance homogeneity** using Levene’s test - **Distribution shape** using IQR similarity

This ensures statistical assumptions are met.

```{r}
# Prepare long-format Day 3 data with group info
day3_long <- data_log %>%
  rownames_to_column("LipidGroup") %>% 
  pivot_longer(-LipidGroup, names_to = "Sample", values_to = "Abundance") %>%
  left_join(coldata %>% dplyr::rename(Sample = `Sample ID`), by = "Sample") %>%
  filter(`Days in incubation` == "3 days") %>%
  mutate(group = `FATP2 status`)

# Normality check per lipid and group
shapiro_results <- day3_long %>%
  group_by(LipidGroup, group) %>%
  summarise(shapiro_p = tryCatch(shapiro.test(Abundance)$p.value, error = function(e) NA_real_), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = shapiro_p, names_prefix = "shapiro_") %>%
  mutate(normal_KO = shapiro_KO > 0.05, normal_WT = shapiro_WT > 0.05)

# Equal variance check
levene_results <- day3_long %>%
  group_by(LipidGroup) %>%
  summarise(levene_p = leveneTest(Abundance ~ group)$`Pr(>F)`[1], .groups = "drop") %>%
  mutate(equal_var = levene_p > 0.05)

# Shape similarity using IQR
iqr_results <- day3_long %>%
  group_by(LipidGroup, group) %>%
  summarise(IQR = IQR(Abundance), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = IQR) %>%
  mutate(similar_shape = abs(WT - KO) / pmax(WT, KO) < 0.3)

# Combine all results
assumptions_summary <- shapiro_results %>%
  left_join(levene_results, by = "LipidGroup") %>%
  left_join(iqr_results, by = "LipidGroup") %>%
  mutate(
    use_ttest = normal_KO & normal_WT & equal_var,
    use_wilcox = !use_ttest & similar_shape,
    skip = is.na(use_ttest) & is.na(use_wilcox)
  )

table(assumptions_summary$use_ttest)
```

------------------------------------------------------------------------

## 🧪 11. Differential Expression Analysis (Welch’s T-test)

For each lipid, we compare KO vs WT on Day 3 using Welch’s t-test (unequal variance). Fold changes and adjusted p-values are calculated.

```{r}
# Prepare clean Day 3 dataset
day3_log <- data_log %>%
  rownames_to_column("LipidGroup") %>%
  pivot_longer(-LipidGroup, names_to = "Sample", values_to = "Abundance") %>%
  left_join(coldata %>% dplyr::rename(Sample = `Sample ID`), by = "Sample") %>%
  filter(`Days in incubation` == "3 days") %>%
  mutate(group = `FATP2 status`, group = factor(group, levels = c("WT", "KO")))

# Compute group-wise means for KO and WT
group_means <- day3_log %>%
  group_by(LipidGroup, group) %>%
  summarise(mean_expr = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = group, values_from = mean_expr, names_prefix = "mean_") %>%
  mutate(
  log10_FC = mean_WT - mean_KO,
  FC = 10 ^ log10_FC,
  log2_FC = log10_FC * log2(10)
)


# Run Welch’s t-test (safe wrapper)
safe_t_test <- safely(~ t.test(Abundance ~ group, data = .x, var.equal = FALSE))

# Run test and merge with FCs
t_test_results <- day3_log %>%
  group_by(LipidGroup) %>%
  nest() %>%
  mutate(
    t_result = map(data, safe_t_test),
    t_output = map(t_result, "result"),
    t_tidy = map(t_output, ~ if (!is.null(.x)) broom::tidy(.x) else NULL)
  ) %>%
  unnest(t_tidy) %>%
  select(LipidGroup, statistic, p.value) %>%
  mutate(p_adj = p.adjust(p.value, method = "fdr")) %>%
  left_join(group_means, by = "LipidGroup") %>%
  select(LipidGroup, mean_WT, mean_KO, FC, log2_FC, statistic, p.value, p_adj)

# Preview results
head(t_test_results)

```

------------------------------------------------------------------------

## 🧬 12. Integrate Lipid Classes and Visualize by Class

We now re-annotate DE results with lipid class information and visualize by class category.

```{r}
lipid_classes <- matrix %>% select(LipidGroup, Class) %>% distinct()
t_test_results <- t_test_results %>%
  left_join(lipid_classes, by = "LipidGroup")


class_plot <- t_test_results %>%
  filter(p_adj <= 0.05) %>%
  ggplot(aes(x = FC, y = Class, color = Class)) +
  geom_point(size = 4) +
  geom_vline(xintercept = 1, linetype = "dashed") +  # FC = 1 = no change
  theme_minimal() +
 labs(
  title = "Significant Differential Lipids by Class (Day 3, padj ≤ 0.05)",
  x = "Fold Change (WT / KO)",
  y = "Lipid Class"
)+
  theme(legend.position = "none")


```
HERE THE ONES THAT ARE SIGNIFICANT
------------------------------------------------------------------------

## 🌋 13. Volcano Plot of Differential Lipids

Highlighting upregulated and downregulated lipids based on padj and log2FC thresholds. Labels added for clarity.

```{r}
library(ggplot2)
library(dplyr)
library(ggrepel)

# Thresholds
padj_threshold <- 0.1
fc_upper <- 1.5
fc_lower <- 0.5

# Categorize differential lipids
volcano_data <- t_test_results %>%
  mutate(
    GeneType = case_when(
      FC >= fc_upper & p_adj <= padj_threshold ~ "up",
      FC <= fc_lower & p_adj <= padj_threshold ~ "down",
      TRUE ~ "ns"
    ),
    Molecule = LipidGroup
  )

# Custom aesthetics
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2.5, "down" = 2.5, "ns" = 1.2)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.05)

# Volcano plot
p <- ggplot(volcano_data, aes(
  x = FC,
  y = -log10(p_adj),
  fill = GeneType,
  size = GeneType,
  alpha = GeneType
)) +
  geom_point(shape = 21, color = "black") +
  ggrepel::geom_text_repel(
    data = subset(volcano_data, GeneType %in% c("up", "down")),
    aes(label = Molecule),
    size = 3,
    max.overlaps = 5,
    box.padding = 0.4,
    point.padding = 0.3,
    segment.size = 0.3,
    show.legend = FALSE
  ) +
  geom_hline(yintercept = -log10(padj_threshold), linetype = "dashed") +
  geom_vline(xintercept = c(fc_lower, fc_upper), linetype = "dashed", color = "gray40") +
  scale_x_continuous(
    breaks = c(0.25, 0.5, 1, 1.5, 2, 4),
    labels = c("0.25", "0.5", "1", "1.5", "2", "4"),
    limits = c(0.25, 4),
    expand = expansion(mult = c(0.01, 0.05))
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = cols, name = "Regulation") +
  scale_size_manual(values = sizes, guide = "none") +
  scale_alpha_manual(values = alphas, guide = "none") +
  labs(
  title = "Volcano Plot (Day 3, Fold Change): WT vs KO",
  subtitle = paste0("Thresholds: FC ≤ ", fc_lower, " or FC ≥ ", fc_upper,
                    " and padj ≤ ", padj_threshold),
  x = "Fold Change (WT / KO)",
  y = expression(-log[10]("Adjusted P-Value"))
)+
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10)
  )

p


```

# Running the function

```{r}
run_de_analysis <- function(data_log, coldata, timepoint_string, padj_threshold) {
  # 1. Prepare long-format data for selected timepoint
  df_long <- data_log %>%
    rownames_to_column("LipidGroup") %>%
    pivot_longer(-LipidGroup, names_to = "Sample", values_to = "Abundance") %>%
    left_join(coldata %>% dplyr::rename(Sample = `Sample ID`), by = "Sample") %>%
    filter(`Days in incubation` == timepoint_string) %>%
    dplyr::mutate(group = factor(`FATP2 status`, levels = c("WT", "KO")))

  # 2. Compute means and log2FC
  group_means <- df_long %>%
    group_by(LipidGroup, group) %>%
    summarise(mean_expr = mean(Abundance, na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = group, values_from = mean_expr, names_prefix = "mean_") %>%
    dplyr::mutate(
  FC = ifelse(mean_WT == 0, NA_real_, mean_KO / mean_WT),
  log2_FC = log2(FC)
)


  # 3. Run Welch’s t-test safely
  safe_t_test <- safely(~ t.test(Abundance ~ group, data = .x, var.equal = FALSE))

  de_results <- df_long %>%
    group_by(LipidGroup) %>%
    nest() %>%
    dplyr::mutate(
      t_result = map(data, safe_t_test),
      t_output = map(t_result, "result"),
      t_tidy = map(t_output, ~ if (!is.null(.x)) broom::tidy(.x) else NULL)
    ) %>%
    unnest(t_tidy) %>%
    select(LipidGroup, statistic, p.value) %>%
    dplyr::mutate(p_adj = p.adjust(p.value, method = "fdr")) %>%
    left_join(group_means, by = "LipidGroup") %>%
    select(LipidGroup, mean_WT, mean_KO, FC, log2_FC, statistic, p.value, p_adj)

  # 4. Add class annotations
  lipid_classes <- matrix %>% select(LipidGroup, Class) %>% distinct()
  de_results <- de_results %>%
    left_join(lipid_classes, by = "LipidGroup")

  # 5. Fold change plot by class
  top_hits <- de_results %>% filter(p_adj < padj_threshold)
  fc_plot <- ggplot(top_hits, aes(x = log2_FC, y = Class, color = Class)) +
    geom_point(size = 4) +
    geom_vline(xintercept = 0, linetype = "dashed") +
    theme_minimal() +
    labs(
      title = paste0("Differential Lipids by Class of KO vs WT— ", timepoint_string, " (padj ≤ ", padj_threshold, ")"),
      x = "log2(Fold Change)",
      y = "Lipid Class"
    ) +
    theme(legend.position = "none")

  # 6. Volcano plot
  volcano_data <- de_results %>%
  filter(!is.na(log2_FC), !is.na(p_adj)) %>%
  mutate(
    GeneType = case_when(
      log2_FC >= 1 & p_adj <= padj_threshold ~ "up",
      log2_FC <= -1 & p_adj <= padj_threshold ~ "down",
      TRUE ~ "ns"
    ),
    Molecule = LipidGroup
  )

 # Only plot if at least one up or down-regulated hit exists
if (any(volcano_data$GeneType %in% c("up", "down"))) {
  volcano_plot <- ggplot(volcano_data, aes(
    x = log2_FC,
    y = -log10(p_adj),
    fill = GeneType,
    size = GeneType,
    alpha = GeneType
  )) +
    geom_point(shape = 21, color = "black") +
    ggrepel::geom_text_repel(
      data = subset(volcano_data, GeneType %in% c("up", "down")),
      aes(label = Molecule),
      size = 3,
      max.overlaps = 15,
      box.padding = 0.4,
      point.padding = 0.3,
      segment.size = 0.3,
      show.legend = FALSE
    ) +
    geom_hline(yintercept = -log10(padj_threshold), linetype = "dashed") +
    geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
    scale_fill_manual(values = cols, name = "Regulation") +
    scale_size_manual(values = sizes, guide = "none") +
    scale_alpha_manual(values = alphas, guide = "none") +
    labs(
      title = paste0("Volcano Plot — of KO vs WT - ", timepoint_string, " (padj ≤ ", padj_threshold, ")"),
      x = "Log2 Fold Change",
      y = "-Log10 Adjusted P-Value"
    ) +
    theme_minimal(base_size = 13)
} else {
  volcano_plot <- ggplot() + 
    theme_void() +
    labs(title = paste0("Volcano Plot — of KO vs WT - ", timepoint_string, ": No significant hits at padj ≤ ", padj_threshold))
}

  # 7. Return results and both plots
  return(list(
    de_results = de_results,
    foldchange_plot = fc_plot,
    volcano_plot = volcano_plot
  ))
}

results_d0 <- run_de_analysis(data_log, coldata, "0 days", padj_threshold = 0.05)
results_d1 <- run_de_analysis(data_log, coldata, "1 day", padj_threshold = 0.1)
results_d3 <- run_de_analysis(data_log, coldata, "3 days", padj_threshold = 0.1)


```
Use 0.5 for Fold change

# Looping to see all days

```{r}
# Define your unique timepoints
timepoints <- unique(coldata$`Days in incubation`)

# Run the analysis for each timepoint
all_de_results <- map(
  timepoints,
  ~ run_de_analysis(data_log, coldata, timepoint_string = .x, padj_threshold = 0.05)
)

# Name the list by timepoint for easy access
names(all_de_results) <- timepoints

```

## Access the results

```{r}
# Get DE results for "3 days"
all_de_results[["3 days"]]$de_results

# Show the fold change plot for "1 day"
all_de_results[["1 day"]]$foldchange_plot

# Print the volcano plot for "0 days"
all_de_results[["0 days"]]$volcano_plot

```
# fix everything here in the x axis

sppecific subspecies and major groups

A 'score' for a Low Medium High

Standard deviation of 30% and see how they look \# Try the plot_trend(two_group)

# Heatmap of different lipids for WT and KO at day 3 including of classes



# Provide day 3 


# PCA withouth Day 0 and 1
```{r}
# Filter out Day 0 and Day 1 samples
filtered_data_log <- data_log %>%
  select(-matches("^D0_|^D1_"))

# 1. Run PCA (transpose because rows = lipids, columns = samples)
pca <- prcomp(t(filtered_data_log), center = TRUE, scale. = TRUE)

# 2. Extract PC1 and PC2 scores per sample
pca_scores <- as.data.frame(pca$x[, 1:2])
pca_scores$SampleID <- rownames(pca_scores)

# 3. Merge PCA scores with metadata
pca_scores_annotated <- pca_scores %>%
  left_join(
    coldata %>% dplyr::rename(SampleID = `Sample ID`),
    by = "SampleID"
  )

# 4. Plot PCA with annotations
pca_plot_day3 <- ggplot(pca_scores_annotated, aes(x = PC1, y = PC2,
                                 color = `Days in incubation`,
                                 shape = `FATP2 status`)) +
  geom_point(size = 3, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "PCA of Lipid Data (Manual, excluding D0 & D1)",
    x = paste0("PC1 (", round(100 * summary(pca)$importance[2, 1], 1), "% variance)"),
    y = paste0("PC2 (", round(100 * summary(pca)$importance[2, 2], 1), "% variance)")
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  )

```



# Saving the plots

```{r}
# Create output directory if it doesn't exist
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Praise/lipidomics/day_3"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# 📊 1. Save PCA plot (without Day 0 and Day 1)
ggsave(
  filename = file.path(output_dir, "PCA_Day3only_noDay0Day1.pdf"),
  plot = pca_plot_day3,  # replace with your actual PCA ggplot object
  width = 11,            # Full-page width (A4 landscape)
  height = 8.5,
  units = "in",
  device = cairo_pdf,    # Better font/text support
  dpi = 600              # High resolution (for raster layers, e.g., labels)
)

# 🌋 2. Save Volcano Plot
ggsave(
  filename = file.path(output_dir, "VolcanoPlot_Day3_FC_KO_vs_WT.pdf"),
  plot = p,              # Volcano plot object
  width = 11,
  height = 8.5,
  units = "in",
  device = cairo_pdf,
  dpi = 600
)

# 🎯 3. Save Differential Lipids by Class Plot
ggsave(
  filename = file.path(output_dir, "DifferentialLipids_ByClass_Day3.pdf"),
  plot = class_plot,     # replace with your class-level ggplot object
  width = 11,
  height = 8.5,
  units = "in",
  device = cairo_pdf,
  dpi = 600
)

# 📦 Required libraries
library(grid)
library(pheatmap)

# 🖼 Generate the heatmap and store the grob object
heatmap_day3 <- pheatmap(
  heatmap_scaled,
  annotation_col = annotation_col,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  fontsize_row = 8,
  fontsize_col = 9,
  main = "Day 3 - Significant Lipid Heatmap (KO vs WT)",
  border_color = NA,
  show_rownames = TRUE,
  show_colnames = F
)

# 💾 Save as high-quality PDF
pdf("/Users/cojulian/Desktop/Tzu_projects/Matt/Praise/lipidomics/day_3/heatmap_day3.pdf",
    width = 10, height = 10, useDingbats = FALSE)  # size in inches
grid::grid.newpage()
grid::grid.draw(heatmap_day3$gtable)
dev.off()


```


## Heatmap

```{r}
library(pheatmap)
library(dplyr)
library(tidyr)

# 1. Extract Day 3 sample names (modify if needed)
day3_samples <- colnames(data_log)[grepl("^D3_", colnames(data_log))]

# 2. Filter significant lipids (e.g., padj ≤ 0.05)
top_lipids <- t_test_results %>%
  filter(p_adj <= 0.05) %>%
  pull(LipidGroup)

# 3. Subset and prepare matrix for heatmap
heatmap_matrix <- data_log[top_lipids, day3_samples]

# Optional: z-score scaling across rows (lipids)
heatmap_scaled <- t(scale(t(heatmap_matrix)))

# 4. Annotate columns (optional)
annotation_col <- coldata %>%
  filter(`Sample ID` %in% day3_samples) %>%
  select(`Sample ID`, `FATP2 status`) %>%
  column_to_rownames("Sample ID")

heatmap_day3 <- pheatmap::pheatmap(
  heatmap_scaled,
  annotation_col = annotation_col,
  clustering_distance_rows = "euclidean",
  clustering_distance_cols = "euclidean",
  clustering_method = "ward.D2",
  color = colorRampPalette(c("navy", "white", "firebrick3"))(100),
  fontsize_row = 9,
  fontsize_col = 9,
  main = "Day 3 - Significant Lipid Heatmap (KO vs WT)",
  border_color = NA,
  show_rownames = TRUE,
  show_colnames = F
)

```

# For all of them KO should be denominator

# The same for each day volcano plot


# Then a huge heatmap

Do a forestplot for lipid classes per day
