---
title: "Isaacs"
format: html
editor: visual
---

```{r}
## Use install.packages() to install these packages:
suppressMessages(library(tidyverse))
suppressMessages(library(janitor))
suppressMessages(library(plotly))
suppressMessages(library(ggrepel))
suppressMessages(library(purrr))
suppressMessages(library(dplyr))

## Use BiocManager::install() to install these packages:
suppressMessages(library(DESeq2))
suppressMessages(library(GeneTonic))
suppressMessages(library(pcaExplorer))
suppressMessages(library(topGO))
suppressMessages(library(tidybulk))
suppressMessages(library(AnnotationDbi))
suppressMessages(library(org.Hs.eg.db))
suppressMessages(library(SummarizedExperiment))
suppressMessages(library(dittoSeq))
suppressMessages(library(msigdbr))
suppressMessages(library(fgsea))

library(tidybulk)
library(ggplot2)
library(ggrepel)
library(SummarizedExperiment)
library(dplyr)
```

# Loading data dictionary

```{r}
data_dict <- read.csv("rnaseq_dictionary_Isaacs.csv")
```

```{r}
# Load packages
library(tidyverse)

# 1. Define the base path to the quant.sf files
quant_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Isaac/multiomics/rnaseq/quant_sf_files_isaac"

# 2. List all quant.sf files recursively
quant_paths <- list.files(quant_dir, pattern = "quant.sf", recursive = TRUE, full.names = TRUE)

# 3. Extract sample folder names like "IV1_S13" from the paths
sample_dirs <- dirname(quant_paths)
sample_ids <- basename(sample_dirs)  # "IV1_S13", etc.

# 4. Extract the Sample code (e.g., "IV1") from the folder name
sample_codes <- gsub("_S[0-9]+", "", sample_ids)

# 5. Build preliminary quant dictionary
quant_dict <- data.frame(
  file_path = quant_paths,
  sample_folder = sample_ids,
  Sample = sample_codes,
  stringsAsFactors = FALSE
)

# 6. Load your existing data_dict (assumed to already be in your environment)
# If it's in a CSV or Excel file, you'd use read_csv() or readxl::read_excel() here

# 7. Join quant_dict with your sample metadata
quant_annotated <- quant_dict %>%
  left_join(data_dict, by = "Sample") %>%
  select(file_path, Sample, Replicate.., Cell.type, Sensitivity.group)

# 8. Print the annotated dictionary
print(quant_annotated)

```

```{r}
# Load the GTF file
gtf <- rtracklayer::import("Homo_sapiens.GRCh38.113.gtf.gz")
gtf_df <- as.data.frame(gtf)

# Create transcript-to-gene table
tx2gene_df <- gtf_df %>%
  dplyr::select(transcript_id, gene_id, gene_name) %>%
  distinct(transcript_id, gene_id, gene_name) %>%
  drop_na()

tx2gene_df
```

```{r}
# Point to your metadata frame
human_coldata <- quant_annotated
human_coldata$files <- human_coldata$file_path  # Add a "files" column for compatibility

# Use first file to initialize matrix dimensions and gene list
tmp <- read.delim(human_coldata$files[1]) %>%
  inner_join(tx2gene_df, by = c("Name" = "transcript_id")) %>%
  group_by(gene_id) %>%
  summarise(NumReads = as.integer(sum(NumReads)))

num.row <- nrow(tmp)
num.col <- length(human_coldata$files)

# Initialize empty matrix
sf.df <- data.frame(matrix(0, nrow = num.row, ncol = num.col))
colnames(sf.df) <- human_coldata$Sample  # Use Sample for column labels
rownames(sf.df) <- tmp$gene_id

```

```{r}
# Loop to fill matrix with counts from each quant.sf
for (i in seq_along(human_coldata$files)) {
  cat("Reading ", human_coldata$files[i], " Now ..............\n")
  
  sf.df[, i] <- read.delim(human_coldata$files[i]) %>%
    inner_join(tx2gene_df, by = c("Name" = "transcript_id")) %>%
    group_by(gene_id) %>%
    summarise(NumReads = as.integer(sum(NumReads))) %>%
    pull(NumReads)
}

# Finalize as matrix
Salmon.matrix <- as.matrix(sf.df)
colnames(Salmon.matrix) <- human_coldata$Sample

# Set rownames to Sample so it aligns with sf.df.data columns
human_coldata <- human_coldata %>% 
  column_to_rownames("Sample")

# Check if all column names in sf.df.data are present in human_coldata rownames
all(colnames(Salmon.matrix) %in% rownames(human_coldata))

```

```{r}
# Convert the matrix to a data frame for tidy manipulation
Salmon.matrix <- as.data.frame(Salmon.matrix)

# Transpose for easier joining and reshaping
Salmon_long <- Salmon.matrix %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "expression")

# Join with metadata to attach Sensitivity.group
Salmon_long <- Salmon_long %>%
  left_join(human_coldata %>% rownames_to_column("Sample"), by = "Sample")

# Define a threshold for minimum expression (e.g., >10)
expression_threshold <- 10
min_samples_per_group <- 2  # e.g., gene must be expressed in at least 2 samples within a group

# For each gene in each group, count how many samples exceed the threshold
gene_group_counts <- Salmon_long %>%
  group_by(gene_id, Sensitivity.group) %>%
  summarise(n_expressed = sum(expression > expression_threshold), .groups = "drop")

# Keep genes that pass the threshold in at least one group
genes_to_keep <- gene_group_counts %>%
  filter(n_expressed >= min_samples_per_group) %>%
  distinct(gene_id)

# Filter original matrix to keep only these genes
sf.df.data <- Salmon.matrix %>%
  rownames_to_column("gene_id") %>%
  filter(gene_id %in% genes_to_keep$gene_id) %>%
  column_to_rownames("gene_id")

# Preview the filtered matrix
head(sf.df.data)

```

# Comparing before and after filtering

```{r}
# 1. Prepare long-format data for unfiltered matrix
Salmon_long_dist_unfiltered <- as.data.frame(Salmon.matrix) %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "expression") %>%
  mutate(stage = "Unfiltered")

# 2. Prepare long-format data for filtered matrix
Salmon_long_dist_filtered <- sf.df.data %>%
  rownames_to_column("gene_id") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "expression") %>%
  mutate(stage = "Filtered")

# 3. Combine both and log-transform expression
combined_expression <- bind_rows(Salmon_long_dist_unfiltered, Salmon_long_dist_filtered) %>%
  mutate(log_expression = log2(expression + 1))  # Avoid log(0)

# 4. Density plot: log2(expression)
ggplot(combined_expression, aes(x = log_expression, fill = stage)) +
  geom_density(alpha = 0.4) +
  labs(
    title = "Distribution of Log2(Expression) Before vs After Filtering",
    subtitle = "Filtering removes low-expression genes across samples",
    x = "log2(Expression + 1)",
    y = "Density",
    fill = "Matrix"
  ) +
  theme_minimal()

# 5. Boxplot: log2(expression)
ggplot(combined_expression, aes(x = stage, y = log_expression, fill = stage)) +
  geom_boxplot(outlier.size = 0.5, alpha = 0.7) +
  labs(
    title = "Boxplot of Log2(Expression) Before vs After Filtering",
    subtitle = "Filtered matrix has higher median expression and tighter range",
    x = "",
    y = "log2(Expression + 1)"
  ) +
  theme_minimal()

# 6. Gene count summary
num_genes_unfiltered <- nrow(Salmon.matrix)
num_genes_filtered <- nrow(sf.df.data)

cat("Number of genes before filtering:", num_genes_unfiltered, "\n")
cat("Number of genes after filtering:", num_genes_filtered, "\n")
cat("Percentage of genes retained:", round((num_genes_filtered / num_genes_unfiltered) * 100, 2), "%\n")
```

# PCA

```{r}
# Load required packages
library(tidybulk)
library(ggplot2)
library(ggrepel)
library(SummarizedExperiment)
library(dplyr)

# 1. Filter out "High" sensitivity samples
filtered_coldata <- human_coldata %>%
  dplyr::filter(Sensitivity.group != "High")

filtered_counts <- sf.df.data[, rownames(filtered_coldata)]

# 2. Create SummarizedExperiment object
human_se <- SummarizedExperiment(
  assays = list(counts = as.matrix(filtered_counts)),
  colData = filtered_coldata
)

# 3. Tidybulk PCA: filter, normalize, reduce
pca_human <- human_se %>%
  tidybulk::keep_abundant() %>%
  tidybulk::scale_abundance() %>%
  tidybulk::reduce_dimensions(method = "PCA", .dims = 15)

# 4. Plot PCA
pca_human %>%
  tidybulk::pivot_sample() %>%
  ggplot(aes(
    x = PC1,
    y = PC2,
    color = Sensitivity.group,
    label = Cell.type  # Label with Cell.type
  )) +
  geom_point(size = 3, shape = 16) +
  ggrepel::geom_text_repel(show.legend = FALSE, max.overlaps = 5) +
  theme_minimal() +
  labs(
    title = "PCA of Human Samples (Excluding High Sensitivity)",
    subtitle = "Colored by Sensitivity Group, labeled by Cell Type",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Sensitivity Group"
  )


```

# Doing QC

```{r}
# Construct SummarizedExperiment for QC
sf.df.se <- SummarizedExperiment(
  assays = list(counts = as.matrix(sf.df.data)),
  colData = human_coldata
)

# === Library Size Distribution ===
library_sizes <- colSums(assay(sf.df.se))

# Create detailed sample names
detailed_names <- paste(human_coldata$Cell.type, human_coldata$Sensitivity.group, human_coldata$Replicate.., sep = "_")


# Plot library size per sample with normal y-axis numbers
ggplot(data.frame(Sample = detailed_names, LibrarySize = library_sizes), aes(x = Sample, y = LibrarySize)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  scale_y_continuous(labels = scales::comma) +  # 👈 This formats y-axis with commas
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 10),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold")
  ) +
  labs(title = "Library Size Distribution", x = "Sample", y = "Library Size (Raw Counts)")

```

## Correlation plot

```{r}
cor_matrix <- cor(assay(sf.df.se))
pheatmap::pheatmap(cor_matrix,
  clustering_distance_rows = "correlation",
  clustering_distance_cols = "correlation",
  main = "Sample Correlation Matrix",
  labels_row = detailed_names,
  labels_col = detailed_names,
  fontsize_row = 9,
  fontsize_col = 9,
  fontsize = 11,
  display_numbers = FALSE,
  color = colorRampPalette(c("navy", "white", "firebrick3"))(50)
)
```

## Dist plot

```{r}
log_counts <- log2(assay(sf.df.se) + 1)

# Melt for ggplot
log_counts_long <- reshape2::melt(log_counts)
colnames(log_counts_long) <- c("Gene", "Sample", "LogExpression")

# Plot
ggplot(log_counts_long, aes(x = LogExpression, color = Sample)) +
  geom_density(size = 1) +
  theme_minimal() +
  labs(
    title = "Log2 Gene Expression Distribution by Sample",
    x = "Log2(Counts + 1)",
    y = "Density"
  ) +
  theme(
    axis.title = element_text(size = 13),
    plot.title = element_text(size = 15, face = "bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 9)
  )
```

# Doing DE

```{r}
# 1. Filter metadata and convert Sensitivity.group to factor
meta_data <- human_coldata %>%
  dplyr::filter(Sensitivity.group %in% c("Medium", "Low")) %>%
  dplyr::mutate(Sensitivity.group = factor(Sensitivity.group))

# 2. Match filtered count matrix to metadata samples
filtered_bulk_matrix <- sf.df.data[, rownames(meta_data)]

# 3. Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(
  countData = as.matrix(filtered_bulk_matrix),
  colData = meta_data,
  design = ~ Sensitivity.group
)

# 4. Run DESeq2
dds <- DESeq(dds)

# 5. Get DE results: Medium vs Low (genes enriched in Medium)
res <- results(dds, contrast = c("Sensitivity.group", "Medium", "Low"), alpha = 0.05)

# 6. Summary of results
summary(res)

# 7. Convert to data frame
de_results_med_vs_low <- as.data.frame(res) %>%
  tibble::rownames_to_column("gene_id")

```

# Volcano plot

```{r}
# Define visual parameters
cols <- c("up" = "#ffad73", "down" = "#26b3ff", "ns" = "grey")
sizes <- c("up" = 2, "down" = 2, "ns" = 1)
alphas <- c("up" = 1, "down" = 1, "ns" = 0.5)

# Prepare DE results for volcano plot
volcano_data <- de_results_med_vs_low %>%
  dplyr::left_join(tx2gene_df %>% dplyr::select(gene_id, gene_name), by = "gene_id") %>%
  dplyr::mutate(GeneSymbol = ifelse(is.na(gene_name), gene_id, gene_name)) %>%
  dplyr::distinct(gene_id, .keep_all = TRUE) %>%
  dplyr::select(GeneSymbol, log2FoldChange, padj) %>%
  dplyr::mutate(FoldChange = 2^log2FoldChange) %>%
  dplyr::mutate(GeneType = dplyr::case_when(
    FoldChange >= 1.5 & padj <= 0.05 ~ "up",
    FoldChange <= (1/1.5) & padj <= 0.05 ~ "down",
    TRUE ~ "ns"
  ))


# Volcano Plot: Medium vs Low Sensitivity
volcano_plot <- ggplot(volcano_data, aes(
    x = log2(FoldChange),
    y = -log10(padj),
    fill = GeneType,
    size = GeneType,
    alpha = GeneType
)) +
    geom_point(shape = 21, colour = "black") +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") +
    geom_vline(xintercept = c(log2(1 / 1.5), log2(1.5)), linetype = "dashed") +
    scale_fill_manual(values = cols) +
    scale_size_manual(values = sizes) +
    scale_alpha_manual(values = alphas) +
    ggrepel::geom_text_repel(
        data = dplyr::filter(volcano_data, GeneType != "ns"),
        aes(
            x = log2(FoldChange),
            y = -log10(padj),
            label = GeneSymbol
        ),
        size = 5,
        color = "darkblue",
        max.overlaps = 10,
        min.segment.length = Inf,
        inherit.aes = FALSE
    ) +
    labs(
        title = "Volcano Plot: Medium vs Low (padj ≤ 0.05, FC ≥ 1.5)",
        x = "log2(Fold Change)",
        y = "-log10(Adjusted P-value)"
    ) +
    theme_minimal(base_family = "Arial") +
    theme(
        panel.background = element_rect(fill = "white", colour = NA),
        plot.background = element_rect(fill = "white", colour = NA),
        legend.position = "bottom",
        plot.title = element_text(size = 20, face = "bold"),
        axis.title.x = element_text(size = 18),
        axis.title.y = element_text(size = 18),
        axis.text = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 16)
    )

# Print the plot
print(volcano_plot)

```

## PA

```{r}
# Load required libraries
library(fgsea)
library(msigdbr)
library(org.Hs.eg.db)
library(dplyr)
library(ggplot2)
library(stringr)

# === 1. Prepare ranked gene list (ENSEMBL -> ENTREZ) ===
ranks <- de_results_med_vs_low %>%
  dplyr::mutate(
    entrez = mapIds(
      org.Hs.eg.db,
      keys = gene_id,
      column = "ENTREZID",
      keytype = "ENSEMBL",
      multiVals = "first"
    )
  ) %>%
  dplyr::filter(!is.na(entrez), !is.na(log2FoldChange)) %>%
  dplyr::arrange(desc(log2FoldChange)) %>%
  dplyr::distinct(entrez, .keep_all = TRUE) %>%
  dplyr::select(entrez, log2FoldChange) %>%
  deframe()

# === 2. Load Hallmark gene sets ===
hallmark_sets <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene) %>%
  split(x = .$entrez_gene, f = .$gs_name)

# === 3. Run fgsea ===
fgseaRes <- fgsea(
  pathways = hallmark_sets,
  stats = ranks,
  nperm = 10000
)

# === 4. Clean and annotate results ===
fgseaResTidy <- fgseaRes %>%
  dplyr::mutate(
    pathway = str_replace_all(pathway, "_", " "),
    NES_direction = ifelse(NES > 0, "Up in Medium", "Up in Low")
  ) %>%
  dplyr::filter(pval < 0.08)

# === 5a. Plot Upregulated pathways (NES > 0) ===
fgseaResTidy %>%
  filter(NES > 0) %>%
  arrange(padj) %>%
  slice_max(order_by = NES, n = 15) %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES)) +
  geom_col(fill = "#61BB6F") +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top Upregulated Hallmark Pathways (Medium vs Low)",
    subtitle = "NES > 0, FDR < 0.08",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)"
  ) +
  theme(axis.text.y = element_text(size = 9))

# === 5b. Plot Downregulated pathways (NES < 0) ===
# Plot top 15 enriched pathways regardless of direction
fgseaResTidy %>%
  arrange(padj) %>%
  slice_max(order_by = abs(NES), n = 15) %>%
  ggplot(aes(x = reorder(pathway, NES), y = NES, fill = NES_direction)) +
  geom_col() +
  scale_fill_manual(values = c("Up in Medium" = "#61BB6F", "Up in Low" = "#EC7012")) +
  coord_flip() +
  theme_minimal(base_size = 13) +
  labs(
    title = "Top Enriched Hallmark Pathways (Medium vs Low)",
    subtitle = "Based on NES, FDR < 0.08",
    x = "Pathway",
    y = "Normalized Enrichment Score (NES)",
    fill = "Direction"
  ) +
  theme(axis.text.y = element_text(size = 9))


```

## GSEA for cell surface

```{r}

# 1. Get GOCC_CELL_SURFACE gene set (ENTREZ IDs)
gocc_cell_surface <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "CC") %>%
  filter(gs_name == "GOCC_CELL_SURFACE") %>%
  distinct(gs_name, entrez_gene) %>%
  split(x = .$entrez_gene, f = .$gs_name)

# 2. Prepare ranked gene list from your DESeq2 result (log2FC by ENTREZ)
gene_ranks <- de_results_med_vs_low %>%
  mutate(entrez = mapIds(org.Hs.eg.db,
                         keys = gene_id,
                         column = "ENTREZID",
                         keytype = "ENSEMBL",
                         multiVals = "first")) %>%
  filter(!is.na(entrez), !is.na(log2FoldChange)) %>%
  distinct(entrez, .keep_all = TRUE) %>%
  arrange(desc(log2FoldChange)) %>%
  pull(log2FoldChange, name = entrez)

# 3. Run fgsea using only the GOCC_CELL_SURFACE gene set
fgsea_surface <- fgsea(
  pathways = gocc_cell_surface,
  stats = gene_ranks,
  nperm = 10000
)

# 4. Plot enrichment curve for this set
cell_surface_plot <- plotEnrichment(
  gocc_cell_surface[[1]],
  stats = gene_ranks
) +
  labs(
    title = "GSEA Enrichment: Cell Surface GO",
    subtitle = "Medium vs Low Sensitivity (RNA-seq)",
    x = "Rank",
    y = "Enrichment Score"
  ) +
  annotate("text", x = -Inf, y = -Inf, hjust = 0, vjust = -0.3,
           label = "Enriched in Medium") +
  annotate("text", x = Inf, y = -Inf, hjust = 1, vjust = -0.3,
           label = "Enriched in Low") +
  theme_minimal(base_size = 13)

```

## Secretome enrichment

```{r}
secretome_file <- readxl::read_xlsx("aaz0274_data_file_s2.xlsx") %>% 
  janitor::clean_names()

# 1. Filter secretome Ensembl IDs (optional: focus on secreted ones only)
secretome_genes <- secretome_file %>%
  filter(grepl("Secreted", annotated_category)) %>%  # Optional: only "Secreted" categories
  pull(ensembl_gene_id) %>%
  unique()

# 2. Convert Ensembl IDs to ENTREZ
secretome_entrez <- mapIds(
  org.Hs.eg.db,
  keys = secretome_genes,
  column = "ENTREZID",
  keytype = "ENSEMBL",
  multiVals = "first"
) %>%
  na.omit() %>%
  unique()

# 3. Build gene set object (same structure as msigdbr sets)
secretome_set <- list("Human Secretome" = secretome_entrez)

# 4. Prepare ranked gene list from your DE results (log2FoldChange named by ENTREZ)
gene_ranks <- de_results_med_vs_low %>%
  mutate(entrez = mapIds(
    org.Hs.eg.db,
    keys = gene_id,
    column = "ENTREZID",
    keytype = "ENSEMBL",
    multiVals = "first"
  )) %>%
  filter(!is.na(entrez), !is.na(log2FoldChange)) %>%
  distinct(entrez, .keep_all = TRUE) %>%
  arrange(desc(log2FoldChange)) %>%
  pull(log2FoldChange, name = entrez)

# 5. Run fgsea using the custom secretome set
fgsea_secretome <- fgsea(
  pathways = secretome_set,
  stats = gene_ranks,
  nperm = 10000
)

# 6. Plot enrichment curve
secretome_plot <- plotEnrichment(
  secretome_set[[1]],
  stats = gene_ranks
) +
  labs(
    title = "GSEA Enrichment: Human Secretome Genes",
    subtitle = "Medium vs Low Sensitivity (RNA-seq)",
    x = "Rank",
    y = "Enrichment Score"
  ) +
  annotate("text", x = -Inf, y = -Inf, hjust = 0, vjust = -0.3,
           label = "Enriched in Medium") +
  annotate("text", x = Inf, y = -Inf, hjust = 1, vjust = -0.3,
           label = "Enriched in Low") +
  theme_minimal(base_size = 13)


```

# Saving plots

```{r}
# Define output directory
output_dir <- "/Users/cojulian/Desktop/Tzu_projects/Matt/Isaac/multiomics/rnaseq/presentation_May6_results"

# 📈 Save Secretome Enrichment Plot (High-res PDF)
ggsave(
  filename = file.path(output_dir, "GSEA_Secretome_Medium_vs_Low.pdf"),
  plot = secretome_plot,
  width = 8,
  height = 6,
  device = cairo_pdf,     # Use Cairo for better font embedding and vector quality
  dpi = 600               # High resolution
)

# 🧬 Save Cell Surface Enrichment Plot (High-res PDF)
ggsave(
  filename = file.path(output_dir, "GSEA_CellSurface_Medium_vs_Low.pdf"),
  plot = cell_surface_plot,
  width = 8,
  height = 6,
  device = cairo_pdf,
  dpi = 600
)

```

## SLC comparison

```{r}
# Load required packages
library(dplyr)
library(ggplot2)

# Extract and process expression for ENSG00000140284 (SLC27A2)
slc27a2_expression <- sf.df.data %>%
  tibble::rownames_to_column("gene_id") %>%
  filter(gene_id == "ENSG00000177455") %>%
  tidyr::pivot_longer(-gene_id, names_to = "Sample", values_to = "counts") %>%
  left_join(human_coldata %>% tibble::rownames_to_column("Sample"), by = "Sample") %>%
  filter(Sensitivity.group %in% c("Low", "Medium")) %>%
  mutate(
    Sensitivity.group = factor(Sensitivity.group, levels = c("Low", "Medium")),
    Sample = factor(Sample, levels = Sample[order(Sensitivity.group)])  # Order samples by group
  )

# Plot per-sample expression
ggplot(slc27a2_expression, aes(x = Sample, y = log2(counts + 1), fill = Sensitivity.group)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = c("Low" = "#26b3ff", "Medium" = "#ffad73")) +
  labs(
    title = "Expression of CD19 (ENSG00000177455) Per Sample",
    x = "Sample",
    y = "log2(Expression + 1)",
    fill = "Sensitivity Group"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )

```

```{r}
# Define gene length (in base pairs)
gene_length <- 7372  # for ENSG00000177455

# Get library sizes (total counts per sample)
library_sizes <- colSums(sf.df.data)

# Prepare table
fpkm_table <- sf.df.data %>%
  tibble::rownames_to_column("gene_id") %>%
  filter(gene_id == "ENSG00000177455") %>%
  pivot_longer(-gene_id, names_to = "Sample", values_to = "counts") %>%
  left_join(human_coldata %>% tibble::rownames_to_column("Sample"), by = "Sample") %>%
  filter(Sensitivity.group %in% c("Low", "Medium")) %>%
  mutate(
    library_size = library_sizes[Sample],
    FPKM = (counts * 1e9) / (gene_length * library_size)
  ) %>%
  dplyr::select(Sample, FPKM, Sensitivity.group)

# View table
fpkm_table

write.csv(fpkm_table, paste0(output_dir,"/cd19_fpkm.csv"))
```

For tomorrow:

# PUT IN FPKM, 3 column table:

Sample name, counts (in fpkm), group 'low/high'

# Secretome and Surfaceome in pdf

------------------------------------------------------------------------

# SCATTER plot of this collapsed using FKPM

# Do the exact same analyses for mass spec

# Plot leading edges
